Supplementary Information -- Morphological convergence as on-line lexical analogy

### 2020-01-29 supplementary_information/si.Rnw
haladunk

# libs
library(tidyverse)
library(xtable)

@@

splive_gcm = test2 %>% 
  mutate(
    c.baseline_gcm_features = scales::rescale(baseline_gcm_features)
  ) %>% 
  filter(word == 'splive') %>% 
  pull(c.baseline_gcm_features) %>% 
  unique %>% 
  round(2)

@@

test2 %>% 
  mutate(
    c.baseline_mgl_features = scales::rescale(baseline_mgl_features)
  ) %>% 
  filter(word == 'splive') %>% 
  select(baseline_mgl_features,c.baseline_mgl_features) %>% 
  distinct()

@@

ITT TARTOK

\section{Residualization of the two categorization models on the baseline data}\label{appendixgcmmglresidualization}

Since the regularity scores of the GCM and the MGL for the nonce verbs are correlated  (r $=$ 0.48), we use residualization to see whether the two models make separate significant contributions to explaining variation in the baseline dataset (Section 6.3 in the main text.)

 The residuals remain significant predictors in both models (M3: MGL score ($\beta$ $=$ 1.27, p $<$ 0.0001), GCM residual ($\beta$ $=$ 1.37, p $<$ 0.0001); M4: GCM score ($\beta$ $=$ 2.29, p $<$ 0.0001), MGL residual ($\beta$ $=$ 0.95, p $<$ 0.0001). 
 Note that while this paper reports values fit to the baseline experiment data, the same pattern of results also holds for models fit to the pre-test data.

\section{Additional information regarding  the role of the GCM and the MGL in the post-test \label{posttestimplementation}}

\subsection{Concordance Indices}
Table \ref{modellingsummary} summarizes the concordance indices between our data and the various categorization models. The individual-GCM improves on the \textsc{Celex}-GCM in terms of explaining the post-test data.   The individual-MGL does not improve on the \textsc{Celex}-MGL.

    \centering
    \begin{tabular}{lllll}
            & training data & method & test data & C (test data)\\
            \hline
\textsc{Celex}-GCM & real verbs & `analogy' & verbs in baseline &{\bf 0.58 }\\
\textsc{Celex}-MGL & real verbs & `rules' & verbs in baseline &{\bf  0.59 }\\ \hline
\textsc{Celex}-GCM & real verbs & `analogy' & verbs in post-test &{\bf  0.6}\\
individual-GCM & real verbs + verbs seen in ESP & `analogy' & verbs in post-test &{\bf  0.68}\\

\textsc{Celex}-MGL & real verbs & `rules' & verbs in post-test &{\bf  0.61}\\
individual-MGL & real verbs + verbs seen in ESP & `rules' & verbs in post-test &{\bf  0.55}\\
\hline
    \end{tabular}
    \caption{Summary of concordance indices between our data and the categorization models. The values above the line show the concordance indices for the baseline data, as reported in Section 6.3 of the main text. The values below the line show the indices for the ESP post-test data.}
    \label{modellingsummary}

 Our concordance indices are calculated using the somers2 function in R's \texttt{Hmisc} package \citep{harrell2014hmisc}.

 Note that the indices in Table \ref{modellingsummary} directly calculate C between the relevant algorithm score (MGL or GCM) and the binary post-test outcome (regular or irregular). As a diagnostic for our regression models, we also calculated concordance indices between model predictions and participants' regular/irregular responses. The concordance indices for multiple regression models are, unsurprisingly, considerably higher than the indices calculated on the basis of a single variable.

 The models in Tables 5 and 6 of the main text both have an index of 0.84, indicating an acceptable level of predictive ability \citep{harrell2013regression}.

 Since the random effects likely account for a large amount of this variance, we calculated the marginal r${^2}$ values for both models: these are 0.27 and 0.22, respectively (the conditional values are 0.45 and 0.47). Conditions and learning models explain some amount of the variation, but the data 
 are noisy.

\subsection{Assessing separate contribution of individual models over \textsc{Celex} models}
The text (Section 6.4 in the main text) explains that we calculate the extra information given by the individual models by subtracting the \textsc{Celex} model prediction from the individual model prediction for each item for each participant, and that we find that the extra information from the individual-GCM  models is predictive over and above the \textsc{Celex}-GCM model.

In Model 5, the extra information from the individual-GCM is a significant predictor of post-test responses  ($\beta$ $=$ 2.61, p $<$ 0.001),  above and beyond the \textsc{Celex}-GCM contribution. In similar Model 6, however, the extra information from the individual-MGL is not a significant predictor of post-test responses ($\beta$ $=$ 0.21, p $=$ 0.23). This means that the individual-GCM has some success in modeling post-test behavior when trained with real English verbs and information from the ESP test, whereas adjustments made to the individual-MGL do not improve predictive power.

@@

@@

In the model in Table 5 in the main text, we elect to mean center continuous variables to counteract collinearity that becomes evident when including the interaction between and {\em lexical typicality} .  Although some debate exists regarding the practice of transforming variables \citep{belsley1991conditioning,echambadi2007mean}, centering variables is common practice to remove nonessential ill conditioning, that is, relationships that inevitably exist between main and product terms \citep{aiken1991multiple,jaccard2003interaction, jaeger2008categorical}. We note that if {\em verb baseline mean} is {\em not} centered, collinearity is high (VIF=15.79) in this model. A non-centered model also yields a significant main effect for {\em lexical typicality}, with increased post-test regularization in the random and reversed conditions. However, we elect to take a more conservative approach by not including a potentially spurious main effect. Note that other than this main effect difference, models with and without centering yield results that are qualitatively indistinguishable; all other main effects and interactions are present in either case. The collinearity for the centered model (Table 5 in the main text) is acceptable, with a maximum VIF score of 2.89.

@@

>>>>

# libs
library(tidyverse)
library(xtable)
library(lme4)

@@

splive_gcm = test2 %>% 
  mutate(
    r_baseline_gcm_features = scales::rescale(baseline_gcm_features)
  ) %>% 
  filter(word == 'splive') %>% 
  pull(r_baseline_gcm_features) %>% 
  unique %>% 
  round(2)

@@

test2 %>% 
  mutate(
    r_baseline_mgl_features = scales::rescale(baseline_mgl_features)
  ) %>% 
  filter(word == 'splive') %>% 
  select(baseline_mgl_features,r_baseline_mgl_features) %>% 
  distinct()

@@

@@

<<main_model1, echo=F,eval=T>>=
test2 = test2 %>% 
  mutate(
    verb_baseline_average = baseline_av - mean(baseline_av, na.rm = T),
    participant_prereg_average = pre_reg_av - mean(pre_reg_av, na.rm = T),
    bot_lexical_typicality = case_when(
        lex_typicality == 'typical' ~ 'typical',
        lex_typicality == 'random' ~ 'random',
        lex_typicality == 'atypical' ~ 'reversed'
      ) %>% 
      factor(levels = c('typical', 'random', 'reversed')),
    bot_regularisation_rate = reg_rate %>% 
      factor(levels = c('nc', '-40%', '+40%'))
  )

##########################################
# from Florian Jaeger

vif.mer <- function (fit) {
  ## adapted from rms::vif
  v <- vcov(fit)
  nam <- names(fixef(fit))
  ## exclude intercepts
  ns <- sum(1 * (nam == "Intercept" | nam == "(Intercept)"))
  if (ns > 0) {
    v <- v[-(1:ns), -(1:ns), drop = FALSE]
    nam <- nam[-(1:ns)] }
  d <- diag(v)^0.5
  v <- diag(solve(v/(d %o% d)))
  names(v) <- nam 
  v 
  }
@

<<main_model2, echo=F,eval=F>>=

lme.fit1 = glmer(resp_post_reg ~ 1 + verb_baseline_average * bot_lexical_typicality + participant_prereg_average + bot_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2, 
  family = binomial (link = logit),
  control=glmerControl(optimizer="bobyqa")
)

save(lme.fit1, file = 'si_fit1.rda')
vif.mer(lme.fit1)
@

<<main_model3, echo=F,eval=T, results=tex>>=
load('si_fit1.rda')
print(xtable(summary(lme.fit1)$coef, digits = 3, label = 'regressiontable1', caption = 'Experimental Data, Post-test Regression Model Summary. Model formula: post-test regular response ~ verb baseline mean x lexical typicality + participant pretest mean + regularization shift + (1 + verb.baseline.mean | participant) + (1 | verb)'))
@

@@

In the model in Table 5 in the main text, repeated here as Table \ref{regressiontable1}, we elect to mean center continuous variables to counteract collinearity that becomes evident when including the interaction between and {\em lexical typicality} .  Although some debate exists regarding the practice of transforming variables \citep{belsley1991conditioning,echambadi2007mean}, centering variables is common practice to remove nonessential ill conditioning, that is, relationships that inevitably exist between main and product terms \citep{aiken1991multiple,jaccard2003interaction, jaeger2008categorical}. We note that if {\em verb baseline mean} is {\em not} centered, collinearity is high (VIF=15.79) in this model. A non-centered model also yields a significant main effect for {\em lexical typicality}, with increased post-test regularization in the random and reversed conditions. However, we elect to take a more conservative approach by not including a potentially spurious main effect. Note that other than this main effect difference, models with and without centering yield results that are qualitatively indistinguishable; all other main effects and interactions are present in either case. The collinearity for the centered model (Table 5 in the main text) is acceptable, with a maximum VIF score of 2.9.

<<Bayes factor, echo=F, eval=F>>=
# random vs atypical

test2b = test2 %>% 
  filter(
    bot_lexical_typicality != 'typical'
  )

fit_power1 = glmer(resp_post_reg ~ 1 + verb_baseline_average * bot_lexical_typicality + participant_prereg_average + bot_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2b, 
  family = binomial ( link = logit ),
  control=glmerControl(optimizer="bobyqa")
  )

fit_power2 = glmer(resp_post_reg ~ 1 + verb_baseline_average + bot_lexical_typicality + participant_prereg_average + bot_regularisation_rate +
    (1 + verb_baseline_average | participant_id) + 
    (1  | word), 
  data = test2b, 
  family = binomial ( link = logit ),
  control=glmerControl(optimizer="bobyqa")
  )

# summary(fit_power1)
# summary(fit_power2)

exp((BIC(fit_power2) - BIC(fit_power1))/2)
@

@@

\section{Fits of the two categorization models on the baseline data}\label{appendixgcmmglresidualization}

We fit the Celex-GCM and the Celex-MGL on the baseline data to see how well they predict regularisation responses by the baseline participants. The Celex-MGL outperforms the Celex-GCM, if slightly. Both models contribute independently to explaining variation in the data.

<<celex_gcm_mgl1, eval=T, echo=F>>=

baseline = baseline %>% 
  mutate(
    r_baseline_mgl_features = scales::rescale(baseline_mgl_features),
    r_baseline_gcm_features = scales::rescale(baseline_gcm_features),
  )

C_baseline_mgl_features = with(baseline, Hmisc::somers2(r_baseline_mgl_features,regular)['C']) %>% 
  round(3)
C_baseline_gcm_features = with(baseline, Hmisc::somers2(r_baseline_gcm_features,regular)['C']) %>% 
  round(3)

mgl_gcm_cor = baseline %>% 
  distinct(word,r_baseline_mgl_features,baseline_gcm_features) %>% 
  summarise(
    cor = cor(r_baseline_mgl_features,baseline_gcm_features)
  ) %>% 
  pull() %>% 
  round(3)

The concordance indices of the Celex-MGL (\Sexpr{C_baseline_mgl_features}) and the Celex-MGL (\Sexpr{C_baseline_gcm_features}) on the baseline data are similar, with some edge retained by the MGL, as in the case of \citep{albright2003rules}. Our concordance indices are calculated using the somers2 function in R's \texttt{Hmisc} package \citep{harrell2014hmisc}.

Since the regularity scores of the GCM and the MGL for the nonce verbs are correlated  (r $=$ \Sexpr{mgl_gcm_cor}), we use residualization to see whether the two models make separate significant contributions to explaining variation in the baseline dataset (Section 6.3 in the main text.)

<<celex_gcm_mgl2, eval=F, echo=F>>=

baseline_words = baseline %>% 
  distinct(word, r_baseline_mgl_features, r_baseline_gcm_features)

celex_m1 = lm(r_baseline_gcm_features ~ 1 + r_baseline_mgl_features, data = baseline_words)
celex_m2 = lm(r_baseline_mgl_features ~ 1 + r_baseline_gcm_features, data = baseline_words)

baseline_words = baseline_words %>% 
  mutate(
    r_baseline_gcm_features_residual = resid(celex_m1),
    r_baseline_mgl_features_residual = resid(celex_m2)
  )

baseline = left_join(baseline, baseline_words, by = c("word", "r_baseline_mgl_features", "r_baseline_gcm_features"))

celex_m3 = glmer(regular ~ 1 + r_baseline_mgl_features + r_baseline_gcm_features_residual + 
                (1 + r_baseline_mgl_features + r_baseline_gcm_features_residual | subject), 
                data = baseline, 
                family = binomial ( link = logit ),
                control=glmerControl(optimizer="bobyqa"))

celex_m4 = glmer(regular ~ 1 + r_baseline_gcm_features + r_baseline_mgl_features_residual + 
                (1 + r_baseline_gcm_features + r_baseline_mgl_features_residual | subject), 
                data = baseline, 
                family = binomial ( link = logit ),
                control=glmerControl(optimizer="bobyqa"))

save(celex_m3, file = 'si_celex_m3.rda')
save(celex_m4, file = 'si_celex_m4.rda')

<<celex_resid1, echo=F,eval=T, results=tex>>=
load('si_celex_m3.rda')
print(xtable(summary(celex_m3)$coef, digits = 3, label = 'resid1', caption = 'MGL scores and GCM residuals, baseline data'))
@

<<celex_resid2, echo=F,eval=T, results=tex>>=
load('si_celex_m4.rda')
print(xtable(summary(celex_m4)$coef, digits = 3, label = 'resid2', caption = 'GCM scores and MGL residuals, baseline data'))
@

The model summaries can be seen in Table \ref{resid1} (M3) and Table \ref{resid2} (M4). The residuals remain significant predictors in both models. Note that while this paper reports values fit to the baseline experiment data, the same pattern of results also holds for models fit to the pre-test data.

\section{Model fits on the post-test data}

We fit the Celex-GCM,the Celex-MGL, as well as the Individual-GCM and the Individual-MGL on the post-test data. Recall that the Celex-based models were trained on verbs in Celex. The individual-based models were trained for each participant, on Celex verbs as well as the co-player's responses in the ESP test. They were tested on the post-test for each participant, individually.

In the paper, we list four techniques of evaluating model fit. We go through these in turn.

\subsection{Comparing concordance indices}

First, we calculated concordance indices for the Celex- and individual-models on the post-test data of the ESP experiment.

<<first, echo=F,eval=T>>=

test2 = test2 %>% 
 mutate(
    r_baseline_mgl_features = scales::rescale(baseline_mgl_features),
    r_baseline_gcm_features = scales::rescale(baseline_gcm_features),
    r_individual_mgl_features = scales::rescale(individual_mgl_features),
    r_individual_gcm_features = scales::rescale(individual_gcm_features)
 )

# C_baseline_mgl_features # celex-mgl on baseline
# C_baseline_gcm_features # celex-gcm on baseline

C_test2_baseline_mgl_features = with(test2, Hmisc::somers2(r_baseline_mgl_features,resp_post_reg)['C']) %>% round(3)
C_test2_baseline_gcm_features = with(test2, Hmisc::somers2(r_baseline_gcm_features,resp_post_reg)['C']) %>% round(3)
C_test2_individual_mgl_features = with(test2, Hmisc::somers2(r_individual_mgl_features,resp_post_reg)['C']) %>% round(3)
C_test2_individual_gcm_features = with(test2, Hmisc::somers2(r_individual_gcm_features,resp_post_reg)['C']) %>% round(3)

Table \ref{modellingsummary} summarizes the concordance indices between our data and the various categorization models. The \textsc{Celex}-MGL, outperforms the \textsc{Celex}-GCM on the post-test data to some extent, much like in the case of our baseline data. The individual-GCM considerably improves on the \textsc{Celex}-GCM in terms of explaining the post-test data. The individual-MGL also improves on the \textsc{Celex}-MGL, but to a lesser extent.

    \centering
    \begin{tabular}{lllll}
            & training data & method & test data & C\\
            \hline
\textsc{Celex}-GCM & real verbs & `analogy' & verbs in baseline &\Sexpr{C_baseline_gcm_features }\\
\textsc{Celex}-MGL & real verbs & `rules' & verbs in baseline &\Sexpr{C_baseline_mgl_features }\\ \hline
\textsc{Celex}-GCM & real verbs & `analogy' & verbs in post-test &\Sexpr{C_test2_baseline_gcm_features}\\
individual-GCM & real verbs + verbs seen in ESP & `analogy' & verbs in post-test &\Sexpr{C_test2_individual_gcm_features}\\

\textsc{Celex}-MGL & real verbs & `rules' & verbs in post-test &\Sexpr{C_test2_baseline_mgl_features}\\
individual-MGL & real verbs + verbs seen in ESP & `rules' & verbs in post-test &\Sexpr{C_test2_individual_mgl_features}\\
\hline
    \end{tabular}
    \caption{Summary of concordance indices between our data and the categorization models. The values above the line show the concordance indices for the baseline data, as reported in Section 6.3 of the main text. The values below the line show the indices for the ESP post-test data.}
    \label{modellingsummary}

The issue here is that a concordance index is not a statistical test, and does not provide a margin of error. The trends are apparent: the individual-GCM is more of an improvement than the individual-MGL. But the strength of these trends is less clear.

\subsection{Extra information provided by the individual model over the \sc{Celex}-model}

Second, we considered the amount of extra variation explained by the individual-model over the \sc{Celex}-model, for the MGL and the GCM, respectively.

Section 6.4 in the main text explains that we calculate the extra information given by the individual models by subtracting the \sc{Celex} model prediction from the individual model prediction for each item for each participant.

<<modeldiff1, echo=F,eval=F>>=

test2 = test2 %>% 
  mutate(
    gcm_score_diff = r_individual_gcm_features - r_baseline_gcm_features,
    mgl_score_diff = r_individual_mgl_features - r_baseline_mgl_features
)

resp_post_reg ~ 1 + verb_baseline_average * bot_lexical_typicality + participant_prereg_average + bot_regularisation_rate

indiv_m5 = glmer(resp_post_reg ~ r_baseline_gcm_features + gcm_score_diff + participant_prereg_average + 
                   (1 + r_baseline_gcm_features + gcm_score_diff | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

save(indiv_m5, file = 'si_indiv_m5.rda')

indiv_m6 = glmer(resp_post_reg ~ r_baseline_mgl_features + mgl_score_diff + participant_prereg_average  + 
                   (1 + r_baseline_mgl_features + mgl_score_diff | participant_id) + (1 | word), 
                 data = test2, 
                 family = binomial (link = logit), 
                 control=glmerControl(optimizer="bobyqa"))

save(indiv_m6, file = 'si_indiv_m6.rda')

<<indiv_diff1, echo=F, eval=T, results=tex>>=

load('si_indiv_m5.rda')
print(xtable(summary(indiv_m5)$coef, digits = 3, label = 'indiv1', caption = 'Celex-GCM and individual score difference, post-test data'))

<<indiv_diff2, echo=F,eval=T, results=tex>>=
load('si_indiv_m6.rda')
print(xtable(summary(indiv_m6)$coef, digits = 3, label = 'indiv2', caption = 'Celex-MGL and individual score difference, post-test data'))
@

The extra information is a significant predictor of post-test responses in both models. However, this effect is much more robust for the GCM than for the MGL, as seen in Table \ref{indiv1} (M5) and Table \ref{indiv2} (M6). The individual-GCM has some success in modeling post-test behavior when trained with real English verbs and information from the ESP test, whereas adjustments made to the individual-MGL do not vastly improve predictive power.

These techniques are indicative of the relative strengths of the individual-MGL and -GCM respective to their \sc{Celex}-counterparts and each other. However, they do not compare them directly using statistical tests.

\subsection{Using combined model predictions to explain variation in participant post-test responses}
